Prehensile Manipulation Planning: Modelling, Algorithms and Implementation

### 2021-10-15 hpp.tex
Last corrections before submission.

Today, robots in industrial manufacturing are mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. However, automating a task with some variability is very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. A good example of this difficulty is
the Amazon picking challenge \cite{brock-2018}. The work described in this
paper is a small step towards simplifying industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or unknown
obstacles. The work only covers motion planning -- and
more accurately manipulation planning. The integration into a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

Therefore this \paper describes a software platform called {Humanoid Path
  Planner} tailored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  nonlinear constraints,
\item an original solver for nonlinear constraints that can handle implicit and explicit constraints,
\item a manipulation planning algorithm that tackles a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state-of-the-art development tools and methods.
\item a docker image of the aforementioned software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found at \href{https://humanoid-path-planner.github.io/hpp-doc}{https://humanoid-path-planner.github.io/hpp-doc}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces nonlinear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

Each \partie is implemented by one or several software packages. For some values that need to be computed, instead of providing formulas, we sometimes provide a link to the C++ or python implementation.

@@

Some systems are subject to nonlinear constraints. These nonlinear
constraints define sub-manifolds of the configuration space the robot
must stay on. For example, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping the same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another method consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} whenever this can be done. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on nonlinear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference lies in the fact that the nodes built on the
tangent space are not immediately projected onto the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} put forward a general framework to plan task-constrained
motions in the presence of moving obstacles. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with nonlinear constraints.

@@

Manipulation planning is a particular instance of path planning where
some objects are moved by robots.
Although several instances of the
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is subject to
nonlinear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable pose of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, where each object is grasped at the most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{LozKae2014}, \cite{18-toussaint-RSS}.

@@

A \textit{kinematic chain} is a tree of \textit{joints} where each
joint represents the mobility of a rigid-body link with respect to
another link or to the world reference frame.
A configuration space called the \textit{joint configuration space} is
associated to each joint.
The most common joints with their respective configuration spaces are
\begin{itemize}
\item linear \textit{translation} with configuration space $\reals$,
\item bounded \textit{rotation} with configuration space  $\reals$,
\item unbounded \textit{rotation} with configuration space  $SO(2)$,
\item \textit{planar} joint with configuration space $SE(2)$,
\item \textit{freeflyer} joint with configuration space $SE(3)$.
\end{itemize}
$SO(n)$ and $SE(n)$ respectively stand for \textit{special orthogonal group} and \textit{special Euclidean group}. They represent the group of rotations and the group of rigid-body transformations in $\reals^n$.

@@

The joint configuration spaces enumerated in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~{\color{blue}Murray \textit{et al}}~\cite[Appendix A]{LMS94} for a thorough definition of Lie groups. Here we detail only properties that are useful for the following developments.

@@

\subsubsection{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are a linear and angular velocities at the origin, $(\mathbf{v},\omega)$ is \textit{transported} to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity \textit{transported} to $g$ corresponds to linear and angular velocities $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to screw motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are (or are isomorphic to) subgroups of $SE(3)$ and follow the same geometrical interpretation.

@@

\section{nonlinear constraints and solvers}\label{sec:constraints}

@@

\subsection{nonlinear constraints}

\begin{definition}\textbf{nonlinear constraint.} A nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non-zero right hand side for the same function $h$. We define for that parameterized nonlinear constraints.
\begin{definition}\textbf{Parameterized nonlinear constraint.}
  A parameterized nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
{\color{blue} Piece-wise} differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

This operator is represented by a matrix with $nv_2$ lines and $nv_1$ columns, where $nv_1$ and $nv_2$ are the dimensions of the tangent spaces of respectively $\liegroup_1$ and $\liegroup_2$.

\subsection{Newton based solver}

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the sub-manifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}{\color{blue}\ominus}\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i{\color{blue}\ominus}\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{who has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications, in which robots manipulate objects, once an object is grasped, the position of the object can be explicitly computed from the configuration of the robot. In this case, some configuration variables of the system depend on other configuration variables:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf_{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.

@@

Similarly, if
\begin{itemize}
\item $m$ and $n$ are two integers,
\item $M$ and $N$ are two subsets of respectively $I_{m}$ and $I_{n}$,
\item $J$ is a matrix with $m$ rows and $n$ columns,
\end{itemize}
we denote by
\begin{equation}\label{eq:sub-matrix}
  J_{M,N}
\end{equation}
the matrix of size $|M| \times |N|$ obtained by extracting the rows of $J$ of indices in $M$ and the columns of $J$ with indices in $N$.

@@

To optimize constraint resolution, we perform variable substitution whenever possible in order to reduce both the number of variables and the dimension of the resulting implicit constraint. Here we describe the method first published in~{\color{blue}Mirabel \textit{et al}}~\cite{MirLam2018}. Unlike in the former paper, the description we give in Algorithm~\ref{alg:ExplicitConstraintSet} is closer to the real implementation. Some links to the source code are indeed provided in the algorithm description.
\begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{initializeSolver}{}
    \State{$explicit$} $\gets$ empty vector of explicit constraints
    \State{$nc \gets 0$}
    \State{$args \gets$ array of size $nq$ filled with -1}
    \EndProcedure
    \Function{add}{$E=(in, out, f)$}
    \If{$args_{out}$ contains an element $\geq 0$}
    \State{return failure}
    \EndIf
    \State{queue $idxArg \gets$ elements of $in$}
    \While{$idxArg$ \textbf{not} empty}
    \State{$iArg \gets idxArg$ first element}
    \State{remove $idxArgs$ first element}
    \If{$iArg\in out$}
    \State{return failure}
    \EndIf
    \If{$args[iArg] == -1$}
    \State{continue}
    \Else
    \State{push $explicit[args[iArg]].in$ elements into $idxArg$}
    \EndIf
    \EndWhile
    \State{fill $args_{out}$ with $nc$}
    \State{explicit.add($E$); $nc\gets nc+1$}
    \State{computeOrder()}
    \State{\Return{success}}
    \EndFunction
  \end{algorithmic}
  \caption{Insertion of an explicit constraint in the solver. Line 1 is called once at initialization of the solver. $explicit$ is a vector that stores the constraints that are successfully added to the solver. $nc$ is the size of the latter. $args$ is an array that stores for each configuration variable, the index in $explicit$ of the constraint that computes this configuration variable, -1 if no constraint computes the index. Procedure \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L144}{ADD} tests whether explicit constraint $E$ is compatible with the previously inserted constraints. Line 6 checks whether any output variable of $E$ is already computed by a previous explicit constraint. If so the procedure returns failure and $E$ is not inserted. The Loop at line 9 recursively checks that any element of $out$ is not an input variable of a previously inserted constraint. If the loop ends without returning failure,  line 18 stores that elements of $out$ are computed by $E$ and $E$ is inserted in the vector of constraints. Function \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L301}{computeOrder} at line 20 recursively computes the order in which the explicit constraints are evaluated, following the rule that the input of a constraint should be evaluated before the output.}
  \label{alg:ExplicitConstraintSet}
\end{algorithm}
Once several compatible explicit constraints have been inserted in the solver,
they behave as a single constraint. For example, if $\conf=(\conf_1,\conf_2,\conf_3)$,
$$
  \left\{\begin{array}{lcl}\conf_1 &=& f_1(\conf_2)\\
  \conf_2 &=& f_2(\conf_3)\end{array}\right. \mbox{ becomes }
  \left[\begin{array}{cc}\conf_1\\ \conf_2\end{array}\right] =
    \left[\begin{array}{cc}f_1(f_2(\conf_3))\\ f_2(\conf_3)\end{array}\right],
$$
and $f_2$ should be evaluated before $f_1$.

@@

Now that we are able to project configurations onto sub-manifolds defined by numerical constraints -- up to some numerical threshold, we need to define paths on such sub-manifolds. The usual way of doing so is by discretizing the path and projecting each sample configuration. The shortcoming is that it requires choosing a discretization step at path construction thus losing the continuous information of the path.

Instead, we propose an alternative architecture where paths store the constraints they are subject to and apply the constraints at path evaluation {\color{blue}(\textit{i.e.} when computing the configuration at a given parameter)}. Let $P\in C^1([0,T],\CS)$ be a path without constraint defined on an interval $[0,T]$, and $\mathbf{proj}$ a projector onto a sub-manifold defined by numerical constraints (\textit{i.e.} an instance of\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}).

@@

The previous sections have presented how we model kinematic chains, configurations and velocities for a given robotic system, how configurations and paths can be projected onto a sub-manifold of the configuration space defined by numerical constraints.

@@

\subsection{Constraint Graph}\label{subsec:constraint graph}

>>>>

Today, robots in industrial manufacturing are mostly programmed by
hand. They repeat the same motion thousands of times with great
accuracy. However, automating a task with some variability is very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. A good example of this difficulty is
the Amazon picking challenge \cite{brock-2018}. The work described in this
paper is a small step towards simplifying industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or unknown
obstacles. The work only covers motion planning -- and
more accurately manipulation planning. The integration into a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

Therefore this \paper describes a software platform called {Humanoid Path
  Planner} tailored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  nonlinear constraints,
\item an original solver for nonlinear constraints that can handle implicit and explicit constraints,
\item a manipulation planning algorithm that tackles a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above, following state-of-the-art development tools and methods.
\item a docker image of the aforementioned software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found at \href{https://humanoid-path-planner.github.io/hpp-doc}{https://humanoid-path-planner.github.io/hpp-doc}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only mentioned in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results for several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces nonlinear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \Partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

Each \partie is implemented by one or several software packages. For some values that need to be computed, rather than providing formulas, we sometimes give a link to the C++ or python implementation.

@@

Some systems are subject to nonlinear constraints. These
constraints define submanifolds of the configuration space the robot
must stay on. For example, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping the same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another method consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} whenever this can be done. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on nonlinear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference lies in the fact that the nodes built on the
tangent space are not immediately projected onto the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} put forward a general framework to plan task-constrained
motions in the presence of moving obstacles. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with nonlinear constraints.

@@

Manipulation planning is a particular instance of path planning where
some objects are moved by robots.
Although several instances of the
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is subject to
nonlinear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
submanifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable pose of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, where each object is grasped at the most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mainly rely on two-level methods
composed of a symbolic task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{LozKae2014}, \cite{18-toussaint-RSS}.

@@

A \textit{kinematic chain} is a tree of \textit{joints} where each
joint represents the mobility of a rigid-body link with respect to
another link or to the world reference frame.
A configuration space called the \textit{joint configuration space} is
associated to each joint.
The most common joints with their respective configuration spaces are
\begin{itemize}
\item linear \textit{translation} with configuration space $\reals$,
\item bounded \textit{rotation} with configuration space  $\reals$,
\item unbounded \textit{rotation} with configuration space  $SO(2)$,
\item \textit{planar} joint with configuration space $SE(2)$,
\item \textit{freeflyer} joint with configuration space $SE(3)$.
\end{itemize}
$SO(n)$ and $SE(n)$ stand for \textit{special orthogonal group} and \textit{special Euclidean group} respectively. They represent the group of rotations and the group of rigid-body transformations in $\reals^n$.

@@

The joint configuration spaces listed in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~{\color{blue}Murray \textit{et al}}~\cite[Appendix A]{LMS94} for a thorough definition of Lie groups. Here we detail only those properties that are useful for the following developments.

@@

\subsubsection{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are linear and angular velocities at the origin, $(\mathbf{v},\omega)$ is \textit{transported} to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity \textit{transported} to $g$ corresponds to linear and angular velocities $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to screw motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are subgroups of $SE(3)$ and follow the same geometrical interpretation.

@@

\section{Nonlinear constraints and solvers}\label{sec:constraints}

@@

\subsection{Nonlinear constraints}

\begin{definition}\textbf{nonlinear constraint.} A nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non-zero right hand side for the same function $h$. For that we define parameterized nonlinear constraints.
\begin{definition}\textbf{Parameterized nonlinear constraint.}
  A parameterized nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
{\color{blue} Piece-wise} differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

This operator is represented by a matrix with $nv_2$ lines and $nv_1$ columns, where $nv_1$ and $nv_2$ are the dimensions of the tangent spaces of $\liegroup_1$ and $\liegroup_2$ respectively.

\subsection{Newton-based solver}

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the submanifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}{\color{blue}\ominus}\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i{\color{blue}\ominus}\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{who has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications in which robots manipulate objects, once an object is grasped, the position of the object can be explicitly computed from the configuration of the robot. In this case, some configuration variables of the system depend on other configuration variables:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf_{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.

@@

Similarly, if
\begin{itemize}
\item $m$ and $n$ are two integers,
\item $M$ and $N$ are two subsets of $I_{m}$ and $I_{n}$ respectively,
\item $J$ is a matrix with $m$ rows and $n$ columns,
\end{itemize}
we denote by
\begin{equation}\label{eq:sub-matrix}
  J_{M,N}
\end{equation}
the matrix of size $|M| \times |N|$ obtained by extracting the rows of $J$ of indices in $M$ and the columns of $J$ with indices in $N$.

@@

To optimize constraint resolution, we perform variable substitution whenever possible in order to reduce the number of variables as well as the dimension of the resulting implicit constraint. Here we describe the method first published in~{\color{blue}Mirabel \textit{et al}}~\cite{MirLam2018}. Unlike in the former paper, the description we give in Algorithm~\ref{alg:ExplicitConstraintSet} is closer to the real implementation. Some links to the source code are indeed provided in the algorithm description.
\begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{initializeSolver}{}
    \State{$explicit$} $\gets$ empty vector of explicit constraints
    \State{$nc \gets 0$}
    \State{$args \gets$ array of size $nq$ filled with -1}
    \EndProcedure
    \Function{add}{$E=(in, out, f)$}
    \If{$args_{out}$ contains an element $\geq 0$}
    \State{return failure}
    \EndIf
    \State{queue $idxArg \gets$ elements of $in$}
    \While{$idxArg$ \textbf{not} empty}
    \State{$iArg \gets idxArg$ first element}
    \State{remove $idxArgs$ first element}
    \If{$iArg\in out$}
    \State{return failure}
    \EndIf
    \If{$args[iArg] == -1$}
    \State{continue}
    \Else
    \State{push $explicit[args[iArg]].in$ elements into $idxArg$}
    \EndIf
    \EndWhile
    \State{fill $args_{out}$ with $nc$}
    \State{explicit.add($E$); $nc\gets nc+1$}
    \State{computeOrder()}
    \State{\Return{success}}
    \EndFunction
  \end{algorithmic}
  \caption{Insertion of an explicit constraint in the solver. Line 1 is called once at initialization of the solver. $explicit$ is a vector that stores the constraints that are successfully added to the solver. $nc$ is the size of the latter. $args$ is an array that, for each configuration variable stores the index in $explicit$ of the constraint that computes this configuration variable, -1 if no constraint computes the index. Procedure \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L144}{ADD} tests whether explicit constraint $E$ is compatible with the previously inserted constraints. Line 6 checks whether any output variable of $E$ is already computed by a previous explicit constraint. If so the procedure returns failure and $E$ is not inserted. The Loop at line 9 recursively checks that any element of $out$ is not an input variable of a previously inserted constraint. If the loop ends without returning failure,  line 18 stores the information that elements of $out$ are computed by $E$ and $E$ is inserted in the vector of constraints. Function \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L301}{computeOrder} at line 20 recursively computes the order in which the explicit constraints are evaluated, following the rule that the input of a constraint should be evaluated before the output.}
  \label{alg:ExplicitConstraintSet}
\end{algorithm}
Once several compatible explicit constraints have been inserted in the solver,
they behave as a single constraint. For example, if $\conf=(\conf_1,\conf_2,\conf_3)$,
$$
  \left\{\begin{array}{lcl}\conf_1 &=& f_1(\conf_2)\\
  \conf_2 &=& f_2(\conf_3)\end{array}\right. \mbox{ becomes }
  \left[\begin{array}{cc}\conf_1\\ \conf_2\end{array}\right] =
    \left[\begin{array}{cc}f_1(f_2(\conf_3))\\ f_2(\conf_3)\end{array}\right],
$$
and $f_2$ should be evaluated before $f_1$.

@@

Now that we are able to project configurations onto submanifolds defined by numerical constraints -- up to some numerical threshold, we need to define paths on such submanifolds. The usual way of doing so is by discretizing the path and projecting each sample configuration. The shortcoming is that it requires choosing a discretization step at path construction thus losing the continuous information of the path.

Instead, we propose an alternative architecture where paths store the constraints they are subject to and apply the constraints at path evaluation {\color{blue}(\textit{i.e.} when computing the configuration at a given parameter)}. Let $P\in C^1([0,T],\CS)$ be a path without constraint defined on an interval $[0,T]$, and $\mathbf{proj}$ a projector onto a submanifold defined by numerical constraints (\textit{i.e.} an instance of\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}).

@@

The previous sections have presented how we model kinematic chains, configurations and velocities for a given robotic system and how configurations and paths can be projected onto a submanifold of the configuration space defined by numerical constraints.

@@

\subsection{Constraint graph}\label{subsec:constraint graph}
